Large-scale hybrid ad hoc network for mobile platforms: Challenges and Experiences

### 2019-02-11 paper.tex
Added system architecture

\title{Large-scale peer-to-peer network for mobile platforms: Challenges and Experiences\\
%\thanks{DAIS-ITA}
}

@@

\begin{abstract}
Peer-to-peer (p2p) networks and Mobile ad hoc networks (MANET) have
been widely studied. However, a real-world deployment for the masses
has remained elusive. Ever-increasing density of mobile devices,
especially in urban areas, has given rise to new applications of p2p
communication. However, the modern smartphone platforms have limited
support for such communications. Further, the issues of battery life,
range, and security remain unaddressed. A key question then is, what
kinds of applications can the modern mobile platforms support and what
challenges remain? This paper identifies a class of applications and
presents a novel p2p architecture called Mesh Network Alerts (MNA) to
support them.  We describe our experiences in deploying MNA as a
real-world peer-to-peer network to millions of users for relaying
severe weather information along with the challenges faced, and the
approaches for addressing them.
\end{abstract}

@@

\section{Introduction}
Mobile devices with programmable platforms such as android and iOS
have steadily grown over the last decade, surpassing the 2 billion
mark \footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/}. MANETs
have been greatly studied given their decentralized nature and
potential for new applications
\cite{loo-manet-2011,perkins-ad-hoc-2001}. Most of the prior work on
p2p networks has focused on valuable analytical and simulation-based
study of MANET behavior
\cite{zhang-topology-manet-2015,marti-misbehavior-manet-2000,mauve-pos-routing-manet-2001}. Given
the outstanding practical challenges of physical nodes, real-world
implementations have been limited and have not reached mass scale
\cite{kiess-manet-impl-2007}. However, with the growth of smartphones,
large-scale real-world implementations may become feasible. This paper
describes a real-world implementation of a p2p delay-tolerant network,
called Mesh Network Alerts (MNA), for relaying severe weather
information to millions of mobile device users as part of the Weather
Channel app\footnote{https://weather.com/apps/ibm/meshnetworkalerts}
on both android and iOS platforms.

@@

%
\btask
%
\item\label{c:0} No secondary communication infrastructure such as WiFi
  access points to fall back on
\item\label{c:1} Network nodes are mobile, pattern of mobility is not
  predictable
\item\label{c:2} New information may arrive at any time
\item\label{c:3} Trustworthy information is scarce, misinformation and
  rumours are common place
\item\label{c:4}  Small payloads suffice in many cases and information
  stays relevant for a few minutes
\item\label{c:5}  Device battery is a scarce resource, power supply
  for recharging may not be available
\item\label{c:6}  Devices are owned by citizens, deployment of
  special-purpose devices is cost prohibitive
%
\etask
%

The above needs are well-recognized in the industry with several
ambitious attempts to address them, e.g., Google Loon
project\footnote{https://loon.co} and Facebook
Aquilla\footnote{https://en.wikipedia.org/wiki/Facebook\_Aquila},
though with limited impact. Leveraging user mobile devices as peer
nodes for a large-scale deployment has been another theme in the prior
works, e.g., the Serval project
\cite{gardner-stephen-serval-2011}. Serval mesh enables p2p
communication over on-device WiFi radio, but requires root access to
the device via jailbreaking. Although significant leassons have been
learned through these attempts, a mass-scale p2p network for such
applications remains elusive.

A vast majority of the literature has focused on a traditional model
of stateful and reliable networking. Specifically, the nodes maintain
connectivity with peers, routing is optimized with techniques based on
link state or distance vectors
\cite{clausen-olsr-2003,perkins-aodv-2003} focusing on optimizing the
network utilization. Given the application characteristics above, this
paper identifies main practical challenges associated with modern
device platforms and finds novel ways to overcome them.  This leads to
MNA --- a new paradigm in p2p networking that employs connection-less,
delay-tolerant, and zero-routing overhead protocols.  MNA is
implemented on both Android and iOS as an SDK and integrated with the
Weather Channel mobile apps. With extensive experiments and experience
of deploying to almost $10$ million users, MNA represents a way
forward for large-scale p2p networks.

In summary, this paper makes the following key contributions:
\begin{itemize}
\item Identification of a class of applications for p2p and their key
  characteristics
\item A deeper investigation of the practical challenges in supporting
  the above class of applications
\item A novel p2p networking SDK using multiple radio
  channels for modern mobile platforms (Android and iOS)
\item Experimental evaluation and deployment statistics
\end{itemize}

@@

\section{Practical Challenges}

@@

meeting the classes of applications described above. These have been

@@

feedback from the makers of Android and iOS.

@@

Due to \ref{c:2}, even when a user is not interacting with the weather
app, or worse yet, when the device is not being used at all, the
devices must discover peer devices to receive and formward potentially
life-saving weather information. Although modern mobile operating
systems such as Android and iOS offer APIs to discover and advertise
information to peer devices over WiFi and Bluetooth interfaces,
peer-to-peer connections do not work when the same APIs are accessed
while the app is in the background. Prior works, widely document these
challenges and take the approach of having special access on the

@@

scale to mass adoption.

@@

processes on each peer must remain active at all times to be able to
discover new information as soon as it arrives. Further, as there is
no back up infrastructure (\ref{c:0}) and a set of peers in range can
change at any time (\ref{c:1}), each peer is responsible for
constantly forwarding available information to other peers via

@@

Given the heterogeneity of devices and operating system distributions
owned by users (\ref{c:6}), it is challenging to test whether or not

@@

As user devices with MNA advertise on a continuous basis over
well-known radio protocols, it may be possible for a malicious
attacker to listen for such advertisements and reverse-engineer the
message formats and protocols used. Then, the attackers may generate
fake messages and advertise them, e.g., a fake tornado alert. Given a
lack of trusted information in such scenarios (\ref{c:3}),
misinformation campaigns can have disastrous consequences.  In open
decentralized systems, such false messages cannot be distinguished
from the real ones, and MNA will end up propagating them to as many
devices as possible, ``poisoning'' the network. In general, veracity
of such information cannot be independently verified in open
decentralized distributed systems and previous work on peer-to-peer
networks do not address this challenge.

@@

\section{System Architecture}

@@

We overcame these challenges via innovative techniques, without
resorting to hacks that may violate user security or Appstore
guidelines. With our technique, exchange of information happens
without making network connections. As we describe later, this is
achieved by splitting the messages into small enough chunks so that
they can be stuffed into service advertisements themselves and
broadcast over multiple advertisements in a quick succession.

Our approach here is two-pronged. Firstly, via extensive experiments
on a large number of devices, we fine-tune the algorithms governing
the intervals at which discovery and advertisements occur. In a
nutshell, receiving new information during a period causes MNA to be
more aggressive in discovery and advertisement. Similarly, lack of new
information for a period makes the device less aggressive. Secondly,
we allow ``wake up'' messages to be broadcast in the network ahead of
an anticipated severe weather event. When devices receive such
messages, they schedule themselves to remain aggressive during the
specified window of time. Outside of this window, the device can
afford to have long sleep cycles and conserve power. With these
techniques, our testing shows less than 1\% battery consumption per
hour on most device models.

We developed test automation tools and processes such that multiple
devices can be controlled from a single test station and follow
prescribed steps to generate, send, and receives messages to play out
a test scenario. Finally, the framework allows automated analysis of
the observations to determine the test result. This capability was
instrumental in uncovering bugs at a fast pace to meet the aggressive
timeline and avoid the cost of acquisition. Further, the automation
framework is general and can be expressly applied to test other apps
in this fashion.

In our approach, since the origin of weather information is the
Weather Channel service, digital signatures can be attached to all
weather alerts broadcast from the service. The mobile application is
distributed with the corresponding public key so that digital
signatures from the service can be verified. If a message fails such a
verification, it is discarded and not forwarded any further.

@@

>>>>

\title{Large-scale hybrid ad hoc network for mobile platforms: Challenges and Experiences}

@@

\begin{abstract}
Peer-to-peer (p2p) networks and Mobile ad hoc networks (MANET) have
been widely studied. However, a real-world deployment for the masses
has remained elusive. Ever-increasing density of mobile devices,
especially in urban areas, has given rise to new applications of p2p
communication. However, the modern smartphone platforms have limited
support for such communications. Further, the issues of battery life,
range, and trust remain unaddressed. A key question then is, what
kinds of applications can the modern mobile platforms support and what
challenges remain? This paper identifies a class of applications and
presents a novel center-to-peer-to-peer (c2p2p) architecture called
Mesh Network Alerts (MNA) to support them.  We describe our
experiences in deploying MNA as a real-world system to millions of
users for relaying severe weather information along with the
challenges faced, and the approaches for addressing them.
\end{abstract}

@@

\section{Introduction}
Mobile devices with programmable platforms such as android and iOS
have steadily grown over the last decade, surpassing the 2 billion
mark \footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/}. MANETs
have been widely studied given their decentralized nature and
potential for new applications
\cite{loo-manet-2011,perkins-ad-hoc-2001}. Most of the prior work on
p2p networks has focused on analytical and simulation-based study of
MANET behavior
\cite{zhang-topology-manet-2015,marti-misbehavior-manet-2000,mauve-pos-routing-manet-2001}. Given
the outstanding practical challenges in deployoing a large number of
physical nodes, real-world implementations have been limited and have
not reached mass scale \cite{kiess-manet-impl-2007}. However, with the
growth of smartphones, large-scale real-world implementations may
become feasible. This paper describes a real-world implementation of a
p2p delay-tolerant network, called Mesh Network Alerts (MNA), for
relaying severe weather information to millions of mobile device users
as part of the Weather Channel
app\footnote{https://weather.com/apps/ibm/meshnetworkalerts} on both
android and iOS platforms.

@@

%
\btask
%
\item\label{c:0} No communication infrastructure such as WiFi
  access points to fall back on
\item\label{c:1} Network nodes are mobile, pattern of mobility is not
  predictable
\item\label{c:2} New information may arrive at any time
\item\label{c:3} Trustworthy information is scarce, misinformation and
  rumours are common place
\item\label{c:4}  Small payloads suffice in many cases and information
  retains value for a few minutes
\item\label{c:5}  Device battery is a scarce resource, power supply
  for recharging may not be available
\item\label{c:6}  Devices are owned by citizens, deployment of
  special-purpose devices is cost prohibitive
%
\etask
%

The above needs are well-recognized in the industry as well as
academia with several ambitious attempts to address them, e.g., Google
Loon project\footnote{https://loon.co} and Facebook
Aquilla\footnote{https://en.wikipedia.org/wiki/Facebook\_Aquila},
though with limited impact. Leveraging user mobile devices as peer
nodes for a large-scale deployment has been another theme in the prior
works, e.g., the Serval project
\cite{gardner-stephen-serval-2011}. Serval mesh enables p2p
communication over on-device WiFi radio, but requires root access to
the device via jailbreaking. Although significant leassons have been
learned through these attempts, a mass-scale p2p network for such
applications remains elusive.

A vast majority of the literature has focused on a traditional model
of stateful, fully decentralized, reliable networking. Specifically,
the nodes maintain connectivity with peers and routing is optimized
with techniques based on link state or distance vectors
\cite{clausen-olsr-2003,perkins-aodv-2003} focusing on optimizing the
network utilization.

Given the application characteristics above, this paper identifies
main practical challenges associated with modern device platforms and
finds novel ways to overcome them.  This leads to MNA --- a new
paradigm in p2p networking that employs a hybrid, delay-tolerant, and
zero-routing overhead architecture.  Unlike previous MANET
architectures, MNA is a hybrid of a centralized and a decentralized
architecture, called \emph{center-to-peer-to-peer (c2p2p)}. A central service
is leveraged as the trusted source of information while the
information is propagated in a decentralized fashion by peers. Such an
architecture allows MNA to bypass the key issue of trust deficit in
open decentralized systems and scale while retaining the ability of
infrastructure-less communication in many application scenarios.

MNA is implemented on both Android and iOS as an SDK and integrated
with the Weather Channel mobile apps. With extensive experiments and
experience of deploying to almost $10$ million users, MNA represents a
way forward for large-scale p2p networks.

This paper makes the following key contributions:
\begin{itemize}
\item Identification of a class of applications for p2p and their key
  characteristics
\item A deeper investigation of the practical challenges in supporting
  the above class of applications
\item A novel c2p2p architecture implementation using multiple radio
  channels for modern mobile platforms (Android and iOS)
\item Experimental evaluation and deployment statistics
\end{itemize}

@@

\section{Challenges}

@@

supporting the classes of applications described above. These have been

@@

feedback from the developers of Android and iOS.

@@

Due to \ref{c:2}, even when a user is not interacting with an app, or
worse yet, when the device is not being used at all, the devices must
continue to discover peer devices to receive and formward
information. Although modern mobile operating systems such as Android
and iOS offer APIs to discover and advertise information to peer
devices over WiFi and Bluetooth interfaces, peer-to-peer connections
do not work reliably when the same APIs are accessed while the app is
in the background. Further, on Android, each WiFi p2p connection must
be explicitly approved by the device user. Prior works widely document
these challenges and take the approach of having special access on the

@@

scale to mass adoption.\\

@@

on each peer must remain active at all times to be able to discover
new information as soon as it arrives. Further, as there is no back up
infrastructure (\ref{c:0}) and a set of peers in range can change at
any time (\ref{c:1}), each peer is responsible for constantly
forwarding available information to other peers via

@@

Given the heterogeneity of devices owned by users and operating system
distributions (\ref{c:6}), it is challenging to test whether or not

@@

As user devices with MNA advertise on unsecure wireless protocols, it
may be possible for a malicious attacker to listen for such
advertisements and reverse-engineer the message formats and protocols
used. Then, the attackers may generate fake messages and advertise
them, e.g., a fake tornado alert. Given a lack of trusted information
in such scenarios (\ref{c:3}), misinformation campaigns can have
disastrous consequences.  In open decentralized systems, such false
messages cannot be distinguished from the real ones, and MNA will end
up propagating them to as many devices as possible, ``poisoning'' the
network. In general, veracity of such information cannot be
independently verified in open decentralized systems and previous works
on peer-to-peer networks do not address this challenge.

@@

\section{C2P2P System Architecture}

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{figs/arch}}
\caption{c2p2p system architecture}
\label{fig:arch}
\end{figure}

To address (or bypass) the challenges identified above, this paper
proposes c2p2p -- a hybrid architecture that leverages a central
service as the sole source of trusted information as depicted in
Figure~\ref{fig:arch}. An application-specific central service, e.g.,
weather.com backend, originates new information as a push message and
assigns it a unique identifier. Next, the central service produces a
digital signature using its private key and appends it to the message
payload. Finally, header parameters specifying TTL (duration after
which the message expires), peer identifier of the central service,
destination peer identifier list (for unicast) or a special identifier
(for broadcast), and the time of origin are added to the message and
the message is sent to a push notification service for distribution.

Mobile applications integrate MNA as an SDK including the public key
of the corresponding central service. Peers having Internet
connectivity receive the push messages and verify the digital
signature to protect against spread of misinformation.  If the message
has not expired and is destined to other peers, then the receiving
peers start forwarding the message to other peers. Depending on the
protocol being used, such forwarding happens as a unicast or
broadcast.

Since new information can arrive at any time or new peers may come
into proximity, and message transmissions are not reliable, all peers
store unexpired and verified messages and continue to forward them to
proximal peers repeatedly. This implies that peers may receive the
same message more than once but since the messages are globally
unique, duplicates can be identified and ignored.  A peer sending a
message adds its own peer identifier to a list of forwarders appended
to the message header.  A receiving peer thus knows all the peers that
already have the message and stops repeating the message to those
peers, controlling the flooding.  As there are no control messages or
any other overhead for routing, we call this \emph{zero-overhead}
routing.

The following describes how the c2p2p architecture addresses the
challenges identified above along with known limitations. Further, to
realize this architecture on mobile devices, we evaluated a variety of
protocols on both Android and iOS. As described later, WiFi DNS on
Android, Bluetooth Nearby on Android, and Blutooth low energy (BTLE)
on both Android and iOS are the only protocols that were found
effective. Actual techniques for discovery, advertisement, and
connection vary across these protocols and are described next.
%
\subsection{Addressing the challenges}
\label{sec:address}
%
\noindent\textbf{Operating system constraints} MNA addresses these
challenges via innovative techniques, without resorting to hacks that
may violate user security or App-store guidelines. DNS service
discovery is a widely supported protocol for all platforms
\cite{cheshire-dns-sd-2013}.

On Android, WiFi DNS improvises on this standard such that the
exchange of information happens without making network connections at
all. This is achieved by splitting the message payloads into small
enough chunks and stuffing them into service advertisements as txt
records and broadcast over multiple advertisements in a quick
succession. Also, we employ foreground services to keep the discovery
and advertisements active indefinitely.

On iOS, the only protocol that stays reliably active in background
without violating iOS developer guidelines is BTLE. Unlike Android,
iOS applications do not need to request discovery and advertisements
contunually. Once a BTLE Peripheral (role that has information) is
advertised and a Central (role that receives information) requests
discovery, the app is allowed to go to sleep. When a matching central
or a peripheral device is found, iOS wakes up the application app to
handle such an event, even when the application is in the
background. MNA builds on this and leverages acynchronous dispatch
queues to turn each peer into a Central as well as a Peripheral
simultaneously. This architecture allows bidirectional data transfers
without concurrency issues.

\noindent\textbf{Power constraints} Due to indefinite forground
services and continuous discovery and advertisement, this is primarily
an issue on Android. Our approach here is two-pronged. Firstly, via
extensive experiments on a large number of devices, we fine-tune the
algorithms governing the intervals at which discovery and
advertisements occur. In a nutshell, receiving new information during
a period causes MNA to be more aggressive in discovery and
advertisement. Similarly, lack of new information for a period makes
the device less aggressive. Secondly, we allow ``wake up'' messages to
be broadcast in the network ahead of an anticipated severe weather
event. When devices receive such messages, they schedule themselves to
remain aggressive during the specified window of time. Outside of this
window, the device can afford to have long sleep cycles and conserve
power. With these techniques, our testing shows less than 2\% battery
consumption per hour on most device models.

\noindent\textbf{Testing p2p networks}
We developed test automation tools and processes such that multiple
devices can be controlled from a single test station and follow
prescribed steps to generate, send, and receives messages to play out
a test scenario. Finally, the framework allows automated analysis of
the observations to determine the test result. This capability was
instrumental in uncovering bugs at a fast pace to meet the aggressive
timeline and avoid the cost of acquisition. Further, the automation
framework is general and can be expressly applied to test other apps
in this fashion.

\noindent\textbf{Trust in information} In our approach, since the
origin of weather information is the Weather Channel service, digital
signatures can be attached to all weather alerts broadcast from the
service. The mobile application is distributed with the corresponding
public key so that digital signatures from the service can be
verified. If a message fails such a verification, it is discarded and
not forwarded any further.

@@

%
\subsection{Automated test framework}
\label{sec:test}
%
